---
title: "Weevils"
author: "John Godlee"
date: "18/07/2018"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(fig.align = "center")
```

# Preamble

What version of R is being used?

```{r}
print(version)

citation(package = "base", lib.loc = NULL, auto = NULL)
```

Citation of packages being used:


```{r}
citation("nlme")
```

```{r message = FALSE, error = FALSE}
library(ggplot2)
library(viridis)
library(dplyr)
library(gstat)  # variogram() 
library(nlme)
library(pander)
library(MuMIn)
library(lme4)
library(rgdal)
library(psych)
library(car)
library(ggraph)
library(igraph)

```


```{r}
damage <- read.csv("data/damage.csv")
site_loc <- read.csv("data/site_loc.csv")
uk_outline <- readOGR("data/GBR_adm", "GBR_adm1")
europe_outline <-readOGR("data/NUTS_RG_10M_2016_4326_LEVL_0/", 
  "NUTS_RG_10M_2016_4326_LEVL_0")
cg_loc <- data.frame(long = -3.21, lat = 55.86)
```

Match site_location information from `site_loc` with individuals in `damage`, using `site_code` and `population` as a key, respectively


```{r}
damage_full <- left_join(site_loc, damage, by = c("site_code" = "population"))
names(damage_full) <- c("site_name", "seed_zone", "site_code", 
  "big_region", "dec_latitude", "dec_longitude", 
  "site_area_ha", "gsl", "growing_deg_days_c", 
  "feb_mean_temp_c", "jul_mean_temp_c", "loc", 
  "family", "individual", "field_code", 
  "x_coord", "y_coord", "xy_coord", 
  "block", "prev_damage", "curr_damage", "total_damage")

damage_nozero_dam <- damage_full %>%
	filter(prev_damage > 0,
				 curr_damage > 0)

```

# Basic exploratory plots

How many saplings received damage from pine weevils?

```{r}
round((1 - (length(damage_full$site_name) - length(damage_nozero_dam$site_name)) / length(damage_full$seed_zone)) * 100, digits = 1)

```

## Map of population collection locations, grouped by region

```{r}
scotland_outline <- uk_outline[uk_outline$NAME_1 == "Scotland",]
```

```{r}
scotland_fort <- fortify(scotland_outline, region = "NAME_1")
```

```{r}
ggplot() + 
	geom_polygon(aes(x = long, y = lat, group = group), 
	  colour = "black", fill = NA,
	  data = scotland_fort, alpha = 1) + 
	geom_point(aes(x = dec_longitude, y = dec_latitude, 
	  colour = seed_zone, shape = big_region),
	  data = site_loc) + 
  geom_point(aes(x = long, y = lat), shape = 23, fill = "black",
    data = cg_loc) + 
#	geom_text(aes(x = dec_longitude, y = dec_latitude,
#								 label = name),
#						 size = 1.6,
#						 data = site_loc) + 
	theme_classic() + 
	theme(legend.position = "right") + 
	coord_map() + 
	xlim(-7.5, 0) + 
	ylim(54.5, 59.5) + 
	xlab("Longitude") + 
	ylab("Latitude")
```

Map of FOV in Western Europe

```{r}
europe_fort <- fortify(europe_outline, region = "NUTS_ID")

```

```{r}
ggplot() + 
	geom_polygon(aes(x = long, y = lat, group = group), 
	  colour = "black", fill = "#CCCCCC",
	  data = europe_fort, alpha = 1) + 
  geom_polygon(aes(x = long, y = lat, group = group), 
	  colour = NA, fill = "#6B6B6B",
	  data = scotland_fort, alpha = 1) + 
  geom_rect(aes(xmin = -8, xmax = 0, ymin = 55, ymax = 60), 
    colour = "red", fill = NA) +
	theme_void() + 
	theme(legend.position = "right") + 
	coord_map() + 
	xlim(-12, 25) + 
	ylim(35, 70) + 
	xlab("Longitude") + 
	ylab("Latitude")
```

## Climate space of sites

```{r}
ggplot(site_loc, aes(x = feb_mean_temp_c, y = jul_mean_temp_c)) + 
	geom_point(aes(colour = seed_zone, shape = big_region)) + 
	labs(x = "February mean temperature (C)",
			 y = "July mean temperature (C)") +
	theme_classic()
```

```{r}
ggplot(site_loc, aes(x = dec_latitude, y = growing_deg_days_c)) + 
	geom_point(aes(colour = seed_zone, shape = big_region)) + 
  stat_smooth(method = "lm") + 
	labs(x = "Decimal latitude",
			 y = "Growing degree days (c)") +
	theme_classic()
```

Latitude and growing days don't correlate strongly, regardless of whether big_region or seed_zone are incorporated as groups.

```{r}
ggplot(site_loc, aes(x = feb_mean_temp_c, y = growing_deg_days_c)) + 
	geom_point(aes(colour = seed_zone, shape = big_region)) + 
	labs(x = "February mean temperature (C)",
		y = "Growing degree days (c)") +
	theme_classic()
```

```{r}
ggplot(site_loc, aes(x = jul_mean_temp_c, y = growing_deg_days_c)) + 
	geom_point(aes(colour = seed_zone, shape = big_region)) + 
	labs(x = "July mean temperature (C)",
		y = "Growing degree days (c)") +
	theme_classic()
```

Both February and July mean temperatures correlate strongly with growing degree days, which makes sense.


```{r}
site_loc %>%
	group_by(seed_zone) %>%
	summarise(mean_growing_deg_days_c = mean(growing_deg_days_c),
	  sd_growing_deg_days_c = sd(growing_deg_days_c)) %>%
	mutate(seed_zone = factor(seed_zone, 
	  levels = seed_zone[order(.$mean_growing_deg_days_c)])) %>%
	ggplot(aes(x = seed_zone, y = mean_growing_deg_days_c)) + 
	geom_bar(stat = "identity", aes(fill = seed_zone)) + 
  geom_errorbar(aes(x = seed_zone, 
    ymin = (mean_growing_deg_days_c - sd_growing_deg_days_c),
    ymax = (mean_growing_deg_days_c + sd_growing_deg_days_c)),
    width = 0.5) + 
	labs(x = "Seed zone",
		y = "Mean growing degree days (C)") + 
	theme_classic() +
	theme(legend.position = "none")
```

There is a good distribution of growing degree days across the seed_zone sites, ranging from `min(site_loc$growing_deg_days_c)` to `max(site_loc$growing_deg_days_c)`.

## Describing the nested structure of the data

```{r}
damage_full$country <- "scotland"

heirarchy <- damage_full %>% 
  select(country, big_region, seed_zone, site_code) %>%
  distinct()

# transform it to an edge list
edges_level1_2 = heirarchy %>% 
  select(country, big_region) %>% 
  unique %>% 
  rename(from=country, to=big_region)

edges_level2_3 = heirarchy %>% 
  select(big_region, seed_zone) %>% 
  unique %>% 
  rename(from=big_region, to=seed_zone)

edges_level3_4 = heirarchy %>% 
  select(seed_zone, site_code) %>% 
  unique %>% 
  rename(from=seed_zone, to=site_code)

edge_list=rbind(edges_level1_2, edges_level2_3, edges_level3_4)
 
# Create graph object
dam_graph <- graph_from_data_frame( edge_list )

# Add line weights by damage
region_weights <- damage_full %>% 
  group_by(big_region) %>%
  summarise(damage = sum(curr_damage))

seed_zone_weights <- damage_full %>%
  group_by(seed_zone) %>%
  summarise(damage = sum(curr_damage)) %>% 
  mutate(seed_zone = factor(as.character(seed_zone), levels = c("N", "NW", "NC", "EC", "NE", "SW", "SC")))
seed_zone_weights <- seed_zone_weights[order(seed_zone_weights$seed_zone), ]

site_code_weights <- damage_full %>%
  group_by(site_code) %>%
  summarise(damage = sum(curr_damage)) %>%
  mutate(site_code = factor(as.character(site_code), levels = c( "GE", "RD", "SO",
    "BE", "LC", "SD", "AM", "GA", "GC", "AB", "GD", "RM", "AC",
    "BB", "GT", "CG", "CR", "GL", "BW", "CC", "MG")))
site_code_weights <- site_code_weights[order(site_code_weights$site_code), ]
  

E(dam_graph)$weight <- c(region_weights$damage, seed_zone_weights$damage, site_code_weights$damage)

ggraph(dam_graph, layout = 'dendrogram', circular = FALSE) + 
  geom_edge_diagonal(aes(width = weight)) +
  geom_node_point() +
  geom_node_label(position = "identity", aes(label = name)) + 
  theme_void() + 
  annotate("rect", xmin = 0, xmax = 22, ymin = 1.6, ymax = 2.4, fill = "red", alpha = 0.1) +
  annotate("rect", xmin = 0, xmax = 22, ymin = 0.6, ymax = 1.4, fill = "blue", alpha = 0.1) + 
  annotate("rect", xmin = 0, xmax = 22, ymin = -0.2, ymax = 0.4, fill = "green", alpha = 0.1) +
  geom_text(aes(x = 24, y = 2), label = "Region", hjust = "right") + 
  geom_text(aes(x = 24, y = 1), label = "Seed\nzone", hjust = "right") +
  geom_text(aes(x = 24, y = 0), label = "Site", hjust = "right")

```

## Boxplots of lesions by Scot's Pine population

Lesions from this year (2015), log+1 transformed, according to regions of seed collection: 
```{r}
ggplot(damage_full, aes(x = site_code, y = log1p(curr_damage))) + 
	geom_boxplot() + 
	labs(x = "Population",
			 y = "Number of lesions in 2015") +
	theme_classic()
```

Zero inflation makes any patterns hard to see.

Previous lesions only, log+1 transformed:

```{r}
ggplot(damage_full, aes(x = site_code, y = log1p(prev_damage))) + 
  geom_boxplot() + 
	labs(x = "Population",
			 y = "Number of old lesions") +
	theme_classic()
```

All lesions, log+1 transformed:

```{r}
ggplot(damage_full, aes(x = site_code, y = log1p(total_damage))) + 
	geom_boxplot() + 
	labs(x = "Population",
			 y = "Total number of lesions") +
	theme_classic()
```


Removing the zeroes, looking at variation within groups for those trees that were damaged.

```{r}
ggplot(damage_nozero_dam, aes(x = site_code, y = log(curr_damage))) + 
	geom_boxplot() + 
	labs(x = "Population",
			 y = "log(Total number of lesions)") +
	theme_classic()
```

Looking at number of lesions across different groupings:

```{r}
ggplot(damage_nozero_dam, aes(x = big_region, y = log1p(total_damage))) + 
	geom_boxplot() + 
	labs(x = "Region",
			 y = "Total number of lesions") +
	theme_classic()
```

```{r}
ggplot(damage_nozero_dam, aes(x = seed_zone, y = log1p(total_damage))) + 
	geom_boxplot() + 
	labs(x = "Region",
			 y = "Total number of lesions") +
	theme_classic()
```


## Is there a correlation between current damage and previous damage?

```{r}
ggplot(damage_full, aes(x = log1p(prev_damage), y = log1p(curr_damage))) + 
	geom_point(aes(colour = seed_zone, shape = big_region)) + 
	stat_smooth(method = "lm") + 
	theme_classic()
```

Not really, not even if we remove all zeroes:

```{r}
ggplot(filter(damage_full, prev_damage > 0, curr_damage > 0), 
	aes(x = log1p(prev_damage), y = log1p(curr_damage))) + 
	geom_point(aes(colour = seed_zone, shape = big_region)) + 
	stat_smooth(method = "lm") + 
	theme_classic()
```

Testing with spearman's correlations:

```{r}
cor_dam <- cor.test(damage_full$prev_damage, damage_full$curr_damage, method = "spearman")

cor_dam_nozero <- cor.test(damage_nozero_dam$prev_damage, damage_nozero_dam$curr_damage, method = "spearman")

cor_dam

cor_dam_nozero
```

There isn't much of a correlation between current damage and previous damage, even when zeroes are removed. Though the relationship is significant for the correlation including zero, presumably because if a tree wasn't damaged last year, it wouldn't be damaged again this year.

Trying with a binomial correlation of damaged or not damaged:

```{r}
damage_full$prev_damage_binom <- ifelse(damage_full$prev_damage > 0, 1, 0)
damage_full$curr_damage_binom <- ifelse(damage_full$curr_damage > 0, 1, 0)

tetrachoric(table(damage_full$prev_damage_binom,damage_full$curr_damage_binom))
```

Still no real strong positive correlation.

## Exploring spatial patterns in data 

Are the populations randomly placed on the grid?

```{r fig.width=14}
ggplot(damage_full, aes(x = x_coord, y = y_coord)) + 
	geom_point(aes(colour = site_name)) + 
	coord_equal() + 
	theme_classic() + 
		theme(legend.position = "bottom") 
```

It would seem so.

Spatial autocorrelation in total number of lesions:

```{r fig.width=14}
ggplot(damage_full, aes(x = x_coord, y = y_coord)) + 
	geom_point(aes(colour = total_damage, size = total_damage)) + 
	scale_colour_viridis(name = "No. lesions") + 
	scale_size_continuous(guide = FALSE) + 
	coord_equal() + 
	theme_classic() + 
	theme(legend.position = "bottom") 

```

Spatial autocorrelation in fresh lesions (this year):

```{r fig.width=14}
ggplot(damage_full, aes(x = x_coord, y = y_coord)) + 
	geom_point(aes(colour = total_damage, size = curr_damage)) + 
	scale_colour_viridis(name = "No. lesions (2015)") + 
	scale_size_continuous(guide = FALSE) + 
	coord_equal() + 
	theme_classic() + 
	theme(legend.position = "bottom") 
```

The weevil damage seems to be clustered in certain areas, possibly the saplings are interacting with each other, with some attracting the weevils and others becoming infected consequentially. The spatial autocorrelation seems to be less pronounced in this year's lesions than previous years.

# Testing for spatial autocorrelation

```{r}
curr_dam_semivar <- variogram(log1p(curr_damage)~1, data = damage_full, locations = ~x_coord+y_coord)

curr_dam_semivar_fit = fit.variogram(curr_dam_semivar, 
        model = vgm(psill = 1, model = "Exp", range = 0.2))

plot(curr_dam_semivar, curr_dam_semivar_fit)
```

There doesn't seem to be any spatial auto-correlation, maybe a sill at ~20, but that's difficult to say.

Removing zero points:

```{r}
damage_nozero_dam <- damage_full %>%
	filter(curr_damage > 0)
```

```{r}
curr_dam_no_zero_semivar <- variogram(log1p(curr_damage)~1, data = damage_nozero_dam, locations = ~x_coord+y_coord)

curr_dam_no_zero_semivar_fit = fit.variogram(curr_dam_no_zero_semivar, 
        model = vgm(psill = 1, model = "Exp", range = 0.2))

plot(curr_dam_no_zero_semivar, curr_dam_no_zero_semivar_fit)
```

No autocorrelation seen when zeroes are removed, just noise. 

# Look for spatial autocorrelation in model residuals

Basic Generalised Least Squares model and outputs:

```{r}
tot_pop_gls <- gls(log1p(curr_damage) ~ site_name , data = damage_full)
vario2 <- variogram(tot_pop_gls$residuals~1, data = damage_full, 
  locations = ~x_coord + y_coord)

plot(curr_dam_semivar)

summary(tot_pop_gls)
```

Construct lots of models of current damage vs. population (`site_code`) with different spatial correlation structures, then compare with AIC:

```{r}
m1 <- gls(log1p(curr_damage) ~ site_name, correlation = corExp(form = ~x_coord + 
    y_coord, nugget = TRUE), data = damage_full)

m2 <- gls(log1p(curr_damage) ~ site_name, correlation = corGaus(form = ~x_coord + 
    y_coord, nugget = TRUE), data = damage_full)

m2_no_nugget <- gls(log1p(curr_damage) ~ site_name, correlation = corGaus(form = ~x_coord + 
    y_coord, nugget = FALSE), data = damage_full)

m3 <- gls(log1p(curr_damage) ~ site_name, correlation = corSpher(form = ~x_coord + 
    y_coord, nugget = TRUE), data = damage_full)

m4 <- gls(log1p(curr_damage) ~ site_name, correlation = corLin(form = ~x_coord + 
    y_coord, nugget = TRUE), data = damage_full)

m5 <- gls(log1p(curr_damage) ~ site_name, correlation = corRatio(form = ~x_coord + 
    y_coord, nugget = TRUE), data = damage_full)

m_no_cor <- gls(log1p(curr_damage) ~ site_name, data = damage_full)

m_null <- gls(log1p(curr_damage) ~ 1, data = damage_full)

aic_table <- data.frame(AIC(m1, m2, m2_no_nugget, m3, m4, m5, tot_pop_gls, m_null, m_no_cor)) %>%
	mutate(model_name = rownames(.)) %>%
	dplyr::select(-df) %>%
	arrange(AIC) %>%
	dplyr::select(model_name, AIC)
pander(aic_table)
```

`m2` (Gaussian spatial autocorrelation) is the best model according to AIC. There is definitely an effect of spatial autocorrelation as the AIC of `m_no_cor` is the worst fitting model of all and is higher than a null model.

Exploring the best fitted model (`m2`). First look at a semivariogram of the normalised residuals and the fit of the correlation structure on the raw values:

```{r}
vario2_resid <- Variogram(m2, form = ~x_coord + y_coord, resType = "normalized")
vario2_fit <- Variogram(m2, form = ~x_coord + y_coord, resType = "pearson")
plot(vario2_resid, smooth = TRUE, ylim = c(0, 1.2))
plot(vario2_fit, ylim = c(0, 1.2))
```

The fit of the correlation structure isn't great (blue line), I still don't understand the semivariogram which decreases at higher distances, but in reading that doesn't seem to be a big issue.

Next, look at the standardized residuals of the model:

```{r}
plot(m2, which = 1:4)
```

Standardized residuals don't show any pattern with the fitted values, so that model assumption isn't violated.

And next the normality of residuals:

```{r}
qqnorm(m2)
```

This is dodgy because the data comes from a zero inflated count distribution, even if it has been log transformed since.

Finally, look at the model outputs:

```{r}
summary(m2)
anova(m2)
r.squaredLR(m2, null = m_null)
r.squaredLR(m2)
```

No significant effect of population on weevil damage once controlled for spatial structure (p = 0.156) and the best model accounts for only 15.6% of the variation in weevil damage, regardless of whether the null model was specified.

# Adding family level predictors to the model

```{r}
m2_fam_nest <- gls(log1p(curr_damage) ~ family:site_name, correlation = corGaus(form = ~x_coord + 
    y_coord, nugget = TRUE), data = damage_full)
```

The model won't converge as the model has a singular fit. I think this is caused by some nested groups having near zero variance, I can remove these groups. It's probably because they have no lesions across an entire group.

```{r}
m2_fam <- gls(log1p(curr_damage) ~ family, 
							correlation = corGaus(form = ~x_coord + y_coord, nugget = TRUE), 
							data = damage_full)

m1_fam <- gls(log1p(curr_damage) ~ family, 
							correlation = corExp(form = ~x_coord + y_coord, nugget = TRUE), 
							data = damage_full)

m3_fam <- gls(log1p(curr_damage) ~ family, 
							correlation = corSpher(form = ~x_coord + y_coord, nugget = TRUE), 
							data = damage_full)

m4_fam <- gls(log1p(curr_damage) ~ family, 
							correlation = corLin(form = ~x_coord + y_coord, nugget = TRUE), 
							data = damage_full)

m5_fam <- gls(log1p(curr_damage) ~ family, 
							correlation = corRatio(form = ~x_coord + y_coord, nugget = TRUE), 
							data = damage_full)
```

```{r}
aic_table <- data.frame(AIC(m1, m2, m2_fam, m2_no_nugget, m3, m4, m5, tot_pop_gls, m_null, m1_fam, m3_fam, m4_fam, m5_fam, m_no_cor)) %>%
	mutate(model_name = rownames(.)) %>%
	dplyr::select(-df) %>%
	arrange(AIC) %>%
	dplyr::select(model_name, AIC)
pander(aic_table)
```

Currently, the GLS with family nested within population doesn't converge due to singularities, I think this is because some of the smaller groups don't have any lesions, so the variance of the group is zero. If I just include family level effects and compare to the previous population level models, I _still_ don't get any better than the population level Gaussian correlation structure model (`m2`).

```{r}
Anova(m2, type = "II")
Anova(m2_fam, type = "II")
```


# Linear mixed effects models - no correlation structure, family level

```{r}
curr_fam_lmer <- lmer(log1p(curr_damage) ~ family + (1|site_name), data = damage_full)
curr_pop_lmer <- lmer(log1p(curr_damage) ~ site_name + (1|block), data = damage_full)
curr_lmer_rand <- lmer(log1p(curr_damage) ~ (1|site_name), data = damage_full)
curr_lmer_null <- lm(log1p(curr_damage) ~ 1, data = damage_full)
```

```{r}
summary(curr_fam_lmer)

aic_table_lmer <- data.frame(AIC(curr_fam_lmer, curr_lmer_rand, curr_lmer_null, curr_pop_lmer)) %>%
	mutate(model_name = rownames(.)) %>%
	dplyr::select(-df) %>%
	arrange(AIC) %>%
	dplyr::select(model_name, AIC)
pander(aic_table_lmer)
```

The null model is better than any family level mixed effects model, and is better than a mixed model including population. This null result is likely because I am unable to include the spatial autocorrelation structure in these linear models, and this takes up a lot of the variance.

# Using larger population groupings and climate variables

Family level groupings didn't appear to predict weevil damage at all and population (site) level groupings don't predict very much either, so what if I group into larger groups such as `seed_zone` and `big_region`?

```{r}
ggplot(damage_full, aes(x = seed_zone, y = log1p(curr_damage))) + 
	geom_boxplot(aes(fill = seed_zone)) + 
	theme_classic() + 
	theme(legend.position = "none")
```

The following model uses the same auto-correlation structure as the previous models, but uses seed zone rather than family as the categorical predictor.
```{r}
m2_sz <- gls(curr_damage ~ seed_zone, 
							correlation = corGaus(form = ~x_coord + y_coord, nugget = TRUE), 
							data = damage_full)
summary(m2_sz)
Anova(m2_sz)
```

This model uses big_region as the categorical predictor:

```{r}
m2_br <- gls(curr_damage ~ big_region, 
							correlation = corGaus(form = ~x_coord + y_coord, nugget = TRUE), 
							data = damage_full)
summary(m2_br)
Anova(m2_br, type = "II")
```


```{r}
aic_table <- data.frame(AIC(m1, m2, m2_no_nugget, m3, m4, m5, tot_pop_gls, m_null, m1_fam, m2_fam, m3_fam, m4_fam, m5_fam, m2_sz, m2_br)) %>%
	mutate(model_name = rownames(.)) %>%
	dplyr::select(-df) %>%
	arrange(AIC) %>%
	dplyr::select(model_name, AIC)
pander(aic_table)
```

Seed zone accounts for more variance than big_region, but both models are vastly worse at predicting weevil damage than population level models.

# Logistic models

This part will form a sort of hurdle model, where there is one logistic model for the binomial response of whether a sapling has weevils or not, and another for saplings with weevils how much damage they get.

Create a binomial variable:

```{r}
damage_full$has_damage <- ifelse(damage_full$curr_damage > 0, 1, 0)
```

Run a logistic regression with grouping and spatial autocorrelation

```{r}
log_model <- glm(has_damage ~ site_name, family = "binomial", data = damage_full)
log_model_null <- glm(has_damage ~ 1, family = "binomial", data = damage_full)

summary(log_model)

AIC(log_model_null, log_model)

```

Bearing in mind that this will change the sample size of each group. By how much?

```{r}
damage_full %>%
  filter(curr_damage > 0) %>%
  group_by(site_code) %>%
  tally() %>% 
  arrange(n) %>%
  pander()
```

Minimum group size is 6, maximum group size is 15.

```{r}
m2_nozero <- gls(log1p(curr_damage) ~ site_name, correlation = corGaus(form = ~x_coord + 
    y_coord, nugget = TRUE), data = damage_nozero_dam)
```
The model accounts for more variation in damage by site when the zeroes are removed. 

# Interpreting optimal model

The best model had no random effect structure and instead just included a spatial autocorrelation term to account for plot effects and excluded trees with no lesions. Of the trees that had lesions these are the model results:

```{r}
AIC(m2_nozero)
r.squaredLR(m2_nozero)
summary(m2_nozero)
Anova(m2_nozero)
```



